---
title: "fitMod: A Unified Interface for Statistical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fitMod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(alloy)
library(nnet)      # für multinom
library(MASS)      # für polr, glm.nb
library(survival)  # für Surv, coxph
library(pscl)      # für zeroinfl
library(AER)       # für tobit
library(lme4)      # für mixed models
```

`fitMod()` provides a unified interface for fitting and displaying a wide 
range of statistical models. The syntax is always the same:
  
```r
fitMod(formula, data, fitfn = "...")
```

For background on the statistical methods, the 
[UCLA Statistical Methods and Data Analytics](https://stats.oarc.ucla.edu) 
site is an excellent reference - we link to the relevant pages throughout.

---
  
## Datasets
  
The following code creates all datasets used in this vignette. Run once,
then save with `usethis::use_data()`.

```r
library(haven)
library(foreign)

# --- Lahigh: absenteeism in Los Angeles high schools ---
Lahigh <- read.dta("https://stats.idre.ucla.edu/stat/data/lahigh.dta")
Lahigh$gender <- relevel(Lahigh$gender, ref = "male")
usethis::use_data(Lahigh)

# --- Admit: graduate school admissions ---
Admit <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
Admit$rank <- factor(Admit$rank)
usethis::use_data(Admit)

# --- Apt: tobit regression example ---
Apt <- haven::read_dta(
  "https://stats.idre.ucla.edu/stat/stata/dae/tobit.dta"
) |> toBaseR()
usethis::use_data(Apt)

# --- Ologit: ordinal logistic regression ---
Ologit <- haven::read_dta(
  "https://stats.idre.ucla.edu/stat/data/ologit.dta"
) |> toBaseR()
usethis::use_data(Ologit)

# --- IceCream: multinomial logistic regression ---
IceCream <- haven::read_sas(
  "https://stats.idre.ucla.edu/wp-content/uploads/2016/02/mlogit.sas7bdat"
)
names(IceCream) <- tolower(names(IceCream))
IceCream$ice_cream <- relevel(
  factor(IceCream$ice_cream,
         labels = c("chocolate", "vanilla", "strawberry")),
  ref = "vanilla"
)
usethis::use_data(IceCream)

# --- Whas100: Worcester Heart Attack Study ---
Whas100 <- haven::read_dta(
  "https://stats.idre.ucla.edu/stat/examples/asa2/whas100.dta"
) |> toBaseR()
Whas100$agex <- cutAge(Whas100$age, full = FALSE)
usethis::use_data(Whas100)

# --- Fish: zero-inflated count data ---
Fish <- read.csv("https://stats.idre.ucla.edu/stat/data/fish.csv")
Fish <- within(Fish, {
  nofish   <- factor(nofish)
  livebait <- factor(livebait)
  camper   <- factor(camper)
})
usethis::use_data(Fish)

# --- BioChemists: biochemists publication data (from pscl) ---
data("bioChemists", package = "pscl")
BioChemists <- bioChemists
usethis::use_data(BioChemists)
```

---
  
## 1. Continuous outcome
  
### Linear regression (`lm`, `lmrob`)
  
```{r}
fitLm <- fitMod(Fertility ~ ., swiss)
fitLm
```

Robust standard errors via `sandwich`:
  
```{r}
fitLm |> print(vcov = "HC3")
```

Robust regression (MM-estimator):
  
```{r}
fitLmrob <- fitMod(Fertility ~ ., swiss, fitfn = "lmrob")
fitLmrob
```

### Gamma regression (`gamma`)

```{r}
fitGamma <- fitMod(Sepal.Length ~ ., iris, fitfn = "gamma")
fitGamma
```

### Tobit regression (`tobit`)

[UCLA reference](https://stats.oarc.ucla.edu/stata/dae/tobit-analysis/)

```{r}
fitTobit <- fitMod(apt ~ read + math + prog, Apt, fitfn = "tobit")
fitTobit
```

---
  
## 2. Binary outcome
  
### Logistic regression (`logit`)
  
[UCLA reference](https://stats.oarc.ucla.edu/r/dae/logit-regression/)

```{r}
fitLogit <- fitMod(admit ~ gre + gpa + rank, Admit, fitfn = "logit")
fitLogit
```

Odds Ratios:
  
```{r}
fitLogit |> print(output = "or")
```

Robust SE:
  
```{r}
fitLogit |> print(vcov = "HC3")
fitLogit |> print(output = "or", vcov = "HC3")
```

Predicted probabilities:

```{r}
head(predict(fitLogit))
pseudoR2(fitLogit)
```

---
  
## 3. Count outcome
  
### Poisson regression (`poisson`)
  
[UCLA reference](https://stats.oarc.ucla.edu/stata/output/poisson-regression/)

```{r}
fitPois <- fitMod(daysabs ~ mathnce + langnce + gender, 
                  data = Lahigh, fitfn = "poisson")
fitPois
fitPois |> print(output = "irr")
fitPois |> print(vcov = "HC0")
```

### Quasi-Poisson (`quasipoisson`)

```{r}
fitQpois <- fitMod(daysabs ~ mathnce + langnce + gender,
                   data = Lahigh, fitfn = "quasipoisson")
fitQpois
```

### Negative binomial (`negbin`)

[UCLA reference](https://stats.oarc.ucla.edu/sas/output/negative-binomial-regression/)

```{r}
fitNegbin <- fitMod(daysabs ~ mathnce + langnce + gender,
                    data = Lahigh, fitfn = "negbin")
fitNegbin
fitNegbin |> print(vcov = "HC3")
```

### Zero-inflated Poisson (`zeroinfl`)

[UCLA reference](https://stats.oarc.ucla.edu/r/dae/zip/)

```{r}
fitZeroinfl <- fitMod(count ~ child + camper | persons,
                      data = Fish, fitfn = "zeroinfl")
fitZeroinfl
```

---
  
## 4. Ordered / Nominal outcome
  
### Ordinal logistic regression (`polr`)
  
[UCLA reference](https://stats.oarc.ucla.edu/r/faq/ologit-coefficients/)

```{r}
fitPolr <- fitMod(apply ~ pared, data = Ologit, fitfn = "polr")
fitPolr
fitPolr |> print(output = "or")
head(predict(fitPolr, output = "both"))
```

### Multinomial logistic regression (`multinom`)

[UCLA reference](https://stats.oarc.ucla.edu/stata/output/multinomial-logistic-regression/)

```{r}
fitMult <- fitMod(ice_cream ~ video + puzzle + female,
                  data = IceCream, fitfn = "multinom")
fitMult
head(predict(fitMult, output = "both"))
```

---
  
## 5. Survival outcome
  
### Cox proportional hazards (`coxph`)
  
[UCLA reference](https://stats.oarc.ucla.edu/stata/examples/asa2/applied-survival-analysis-by-hosmer-lemeshow-and-maychapter-4-interpretation-of-a-fitted-proportional-hazards-regression-model/)

```{r}
fitCox <- fitMod(Surv(foltime, folstatus) ~ gender + agex,
                 data = Whas100, fitfn = "coxph")
fitCox
head(predict(fitCox))
```

### Parametric survival models (`weibull`, `exponential`, `lognormal`, `loglogistic`)

Parametric survival models use the Accelerated Failure Time (AFT) 
framework. Coefficients are displayed as Time Ratios (`TR = exp(coef)`):
a value greater than 1 indicates longer survival, less than 1 shorter
survival.  The shape parameter determines the hazard function: shape > 1
means increasing hazard, shape < 1 decreasing hazard.

```{r}
fitWeibull <- fitMod(Surv(foltime, folstatus) ~ gender + age + bmi,
                     data = Whas100, fitfn = "weibull")
fitWeibull
fitWeibull |> print(output = "coef")   # log Time Ratios
head(predict(fitWeibull))
```

---

## 6. Mixed models

Mixed models extend standard regression by adding random effects to
account for clustered or longitudinal data.  The random effects
structure is specified in the formula using `lme4` notation:
`(1 | group)` for a random intercept, `(1 + x | group)` for random
intercept and slope.

[UCLA reference — linear mixed models](https://stats.oarc.ucla.edu/r/dae/linear-mixed-effects-models-using-r/)  
[UCLA reference — mixed logistic regression](https://stats.oarc.ucla.edu/r/dae/mixed-effects-logistic-regression/)

### Linear mixed model (`lmMixed`)

```{r}
fitLmm <- fitMod(Reaction ~ Days + (1 | Subject),
                 data = lme4::sleepstudy, fitfn = "lmMixed")
fitLmm
head(predict(fitLmm))
```

### Mixed logistic regression (`logitMixed`)

[UCLA reference](https://stats.oarc.ucla.edu/r/dae/mixed-effects-logistic-regression/)

```{r}
fitLogitMixed <- fitMod(use ~ age + urban + (1 | district),
                        data = alloy::Contraception, fitfn = "logitMixed")
fitLogitMixed
fitLogitMixed |> print(output = "or")
head(predict(fitLogitMixed, output = "both"))
```

### Mixed Poisson regression (`poissonMixed`)

```{r}
fitPoisMixed <- fitMod(incidence ~ period + (1 | herd),
                       data = lme4::cbpp, fitfn = "poissonMixed")
fitPoisMixed
fitPoisMixed |> print(output = "irr")
```

### Mixed negative binomial regression (`negbinMixed`)

```{r}
fitNegbinMixed <- fitMod(incidence ~ period + (1 | herd),
                         data = lme4::cbpp, fitfn = "negbinMixed")
fitNegbinMixed
```

---
  
## 7. Machine learning

All ML models use the same `predict()` interface with 
`output = "prob"`, `"class"`, or `"both"`.

```{r}
fitSvm <- fitMod(ice_cream ~ video + puzzle + female,
                 data = IceCream, fitfn = "svm")
head(predict(fitSvm, output = "both"))
```

```{r}
fitRf <- fitMod(ice_cream ~ video + puzzle + female,
                data = IceCream, fitfn = "randomForest")
head(predict(fitRf, output = "both"))
```

```{r}
fitNnet <- fitMod(ice_cream ~ video + puzzle + female,
                  data = IceCream, fitfn = "nnet")
head(predict(fitNnet, output = "both"))
```

```{r}
fitC5 <- fitMod(ice_cream ~ video + puzzle + female,
                data = IceCream, fitfn = "C5.0")
head(predict(fitC5, output = "both"))
```

```{r}
fitNbayes <- fitMod(ice_cream ~ video + puzzle + female,
                    data = IceCream, fitfn = "naiveBayes")
head(predict(fitNbayes, output = "both"))
```

```{r}
fitLda <- fitMod(ice_cream ~ video + puzzle + female,
                 data = IceCream, fitfn = "lda")
head(predict(fitLda, output = "both"))
```

```{r}
fitGlmnet <- fitMod(ice_cream ~ video + puzzle + female,
                    data = IceCream, fitfn = "glmnet")
head(predict(fitGlmnet, output = "both"))
```

```{r}
fitXgb <- fitMod(ice_cream ~ video + puzzle + female,
                 data = IceCream, fitfn = "xgboost")
head(predict(fitXgb, output = "both"))
```

```{r}
fitRpart <- fitMod(ice_cream ~ video + puzzle + female,
                   data = IceCream, fitfn = "rpart")
head(predict(fitRpart, output = "both"))
```

---
  
## 8. Robust standard errors

For `lm`, `glm`, and `lmrob` models, heteroscedasticity-consistent 
standard errors can be requested via the `vcov` argument (powered by 
the `sandwich` package). `"HC3"` is recommended for small samples.

| Type | Description |
|------|-------------|
| `"HC0"` | White (1980) |
| `"HC1"` | HC0 with df correction |
| `"HC3"` | Recommended for small n |
  
```{r}
fitLm     |> print(vcov = "HC3")
fitPois   |> print(vcov = "HC0")
fitNegbin |> print(vcov = "HC3")
```

---
  
## 9. Original model output

Use `output = "genuine"` to see the original `summary()` output of 
the underlying model object:
  
```{r}
fitLm      |> print(output = "genuine")
fitCox     |> print(output = "genuine")
fitWeibull |> print(output = "genuine")
```
